Brundle Development

### 2018-01-13 Manuscript_Figures.Rmd
Updated Figure
Update Markdown to reflect new Brundle function
Reworked Rmd for Manuscript

```{r results='hide', message=FALSE, warning=FALSE}
library(DiffBind)

```{r, fig.cap="Comparison of simple normalisation strategies employed. MA plots showing the changes in ER binding after 48 hours treatment with 100 nM fulvestrant. Three simple normalisation methods were applied to this data and compared to the raw count data. (A) Raw counts. (B) Reads Per Million (RPM) reads in peaks. (C) RPM aligned reads. (D) RPM total reads. Note that the highlighted peaks remain above zero under all three standard normalisations."}

## MA Plot of H2Av normalization
```{r, fig.cap="MA plots showing ER binding before and after treatment with fulvestrant including matched Dm H2Av spike-in control.} (A) Reads corrected to total aligned reads showed the same off-centre peak density as observed in Figure 1. Putative unchanged ER binding sites are within the red triangle. (B) Overlaying the MA plot combining the changes in chromatin binding of Hs ER (black) and Dm H2Av (blue). Dm peaks overlay the off-centre peak density. (C) Utilising the Dm H2Av binding events as a ground truth for 0-fold change, a linear fit to the log-fold change is generated and the fit is applied to adjust the Hs ER binding events."}
par(mfrow=c(1,3))
# Read consensus peaks
hsconsensus<-readRDS("Rdata/015_SLX-8047_hsconsensus.rds")
dmconsensus<-readRDS("Rdata/015_SLX-8047_dmconsensus.rds")

@@

Figure 3 can be viewed interactively on the [USCS track](http://genome.ucsc.edu/cgi-bin/hgTracks?hgS_doOtherUser=submit&hgS_otherUserName=andrewholding&hgS_otherUserSessionName=Brundle).

## MA Plot of CTCF Peaks

```{r, results='hide', message=FALSE, warning=FALSE, fig.cap="Comparison of the control regions used to normalise ER analysis before and after treatment. Dots highlighted in red are significant (FDR = 0.01). The CTCF peaks used for normalisation show no significant change in the number reads before and after treatment."}

source('package/brundle.R')
filename<-"Rdata/040_SLX-14438_dba_human_ER_CTCF.rda"
load(filename)
jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)

jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames
)

jg.plotDeSeq(jg.controlResultsDeseq,
             p=0.01, 
             title.main="Fold-change in CTCF binding",
             flip=T
)

##H2av With DiffBind

```{r results='hide', message=FALSE, warning=FALSE,   out.width = '49%', out.extra='style="float:left"'}
jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX8047_dm.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX8047_hs.csv"
filename<-"Rdata/029_SLX-8047_dba_human_drosophila.rda"

@@

@@

```{r results='hide', message=FALSE, warning=FALSE, out.width = '49%', out.extra='style="float:left"'}
  dbaExperimentAnalysisCorrected<-dba.analyze(dbaExperimentCorrected)
  dba.plotMA(dbaExperimentAnalysisCorrected)
```

@@

@@

```{r results='hide', message='false', warning='false', fig.cap="Comparison of mean counts in CTCF peaks before and after treatment. If the samples have no systematic bias before and after treatment then the linear fit would be expected to have a gradient of 1. Here, we establish that the gradient is < 1, implying a systematic bias between samples. The read counts in the treated samples peaks are corrected (blue), removing the bias, and resulting in a new gradient of 1."}

@@

@@

```{r results='hide', echo=FALSE,warning=FALSE,message=FALSE,error=FALSE} 
jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX8047_dm.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX8047_consensus.csv"

@@

```{r out.width = '50%', out.extra='style="float:left"'}
a + as.layer(b) 
```

```{r out.width = '50%', out.extra='style="float:left"', fig.cap="Comparison of normalisation methods using consensus peak set. (A) The analysis for the CTCF normalised (blue) and H2Av normalised (green) dataset using an ER consensus peak set of 10,000 peaks were formatted as an MA plot and overlaid. This recovered the low-fold change higher-intensity peaks that were not visible in Figure \ref{fig:ERCTCF}A and both datasets showed a similar distribution. (B) Comparison of fold-change values for individual ER binding sites between two datasets showed that the inclusion of these sites did not appear to affect the correlation (r = 0.77)."}

@@

@@

```{r, results='hide', message=FALSE, warning=FALSE}

@@

jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

```{r, fig.cap="Correlation Plots of Replicate Experiments. (A) Scatter plots showing the correlation between the replicates with the lowest correlation value. This is provided both the control (top) and treatment (bottom) conditions. The plotted condition is highlighted with thick border in tables on the right. Colour represents density, blue = lowest, red = highest. (B) Tables showing the correlation coefficient for been each replicated. "}

@@

##Venn diagrams

```{r  out.width = '25%', out.extra='style="float:left"'}
filename<-"Rdata/040_SLX-14438_dba_human_ER_CTCF.rda"
load(filename)

dba.plotVenn(dbaControl,dbaControl$masks$Fulvestrant, label1="Rep1", label2="Rep2", label3="Rep3", main="CTCF Peaks Fulvestrant")
dba.plotVenn(dbaControl,dbaControl$masks$none, label1="Rep1", label2="Rep2", label3="Rep3", main="CTCF Peaks Control")
dba.plotVenn(dbaExperiment,dbaExperiment$masks$Fulvestrant, label1="Rep1", label2="Rep2", label3="Rep3", main="ER Peaks Fulvestrant")
dba.plotVenn(dbaExperiment,dbaExperiment$masks$none,label1="Rep1", label2="Rep2", label3="Rep3", main="ER Peaks Control")
```

```{r, out.width = '25%', out.extra='style="float:left"'}

filename<-"Rdata/026_SLX-8047_dba_drosophila.rda"
load(filename)
filename<-"Rdata/026_SLX-8047_dba_human.rda"
load(filename)

dba.plotVenn(dba_dm,dba_dm$masks$Fulvestrant, label1="Rep1", label2="Rep2", label3="Rep3",label4="Rep4", main="H2av Peaks Fulvestrant")
dba.plotVenn(dba_dm,dba_dm$masks$none,label1="Rep1", label2="Rep2", label3="Rep3",label4="Rep4", main="H2av Peaks Control")
dba.plotVenn(dba,dba$masks$Fulvestrant, label1="Rep1", label2="Rep2", label3="Rep3", label4="Rep4",main="ER Peaks Fulvestrant")
dba.plotVenn(dba,dba$masks$none,label1="Rep1", label2="Rep2", label3="Rep3",label4="Rep4", main="ER Peaks Control")
```

```{r, out.width = '25%', out.extra='style="float:left"'}
filename<-"Rdata/014_SLX-12998_dba_mouse.rda"
load(filename)

dba.plotVenn(dba,dba$masks$Fulvestrant, label1="Rep1", label2="Rep2", label3="Rep3", label4="Rep4",main="Mouse ER Peaks Fulvestrant")
dba.plotVenn(dba,dba$masks$none,label1="Rep1", label2="Rep2", label3="Rep3",label4="Rep4", main="Mouse ER Peaks Control")

filename<-"Rdata/015_SLX-12998_dba_human.rda"
load(filename)

dba.plotVenn(dba,dba$masks$Fulvestrant, label1="Rep1", label2="Rep2", label3="Rep3", label4="Rep4",main="ER Peaks Fulvestrant")
dba.plotVenn(dba,dba$masks$none,label1="Rep1", label2="Rep2", label3="Rep3",label4="Rep4", main="ER Peaks Control")
```

Venn diagrams showing peak overlap between replicates. (A) Peak overlap between samples for CTCF internal control. (B) Peak overlap for murine chromatin spike-in samples. (C) Peak overlap for Drosophila chromatin spike-in. As expected, samples treated with fulvestrant showed fewer ER peaks and a higher proportion of peaks unique to individual samples. This is indicative in a loss of ER binding leading and the resulting decrease in single to noise ratio. As we use a statistical analysis and model the variance of the peaks between replicates, false positives in the treated condition will be removed at the point of differential binding analysis.

@@

```{r}

@@

```{r, fig.cap="Distribution of reads for Mm chromatin spike-in normalisation strategy. Comparison of murine chromatin between samples showed no systematic bias in the sample preparation. Bar plots (left axis) represent the fraction of total aligned reads. The dot plot represents the total aligned reads (right axis) for each sample."}
filename<-"Rdata/012_SLX-12998_nrreads.rda"
load(filename)

@@

```{r}
par(mfrow=c(1,3))

@@

```{r,results='hide', message=FALSE, warning=FALSE,   out.width = '33%', out.extra='style="float:left"'}
filename<-"Rdata/015_SLX-14229_dbacounts_human.rda"
load(filename)
plot(dbacounts)

@@

@@

```{r, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}

@@

```{r, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}

@@

##Comparison of mormalisation DiffBind plots

```{r,results='hide', message=FALSE, warning=FALSE,   out.width = '32%', out.extra='style="float:left"'}
jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX14438_hs_CTCF_DBA.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX14438_hs_ER_DBA.csv"

@@

jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

```{r results='hide', message=FALSE, warning=FALSE, fig.cap="Comparison of fold-change of ER binding after both xenogenic and cross-normalisation. Pearson's correlation between the two methods is 0.992 (3sf, p-value tending to 0). Deviation of data points from parity is a result the integer nature of read counts, nonetheless this effect is is very small as demonstrated correlation coefficient between the two datasets."}
filename<-"Rdata/047_SLX-14438_dba_human_ER_CTCF.rda"
load(filename)

@@

@@

#Get sample names for conditions
jg.untreatedNames <- names(jg.ERCountsTreated)
jg.treatedNames   <- names(jg.ERCountsUntreated)

@@

#Get sample names for conditions
jg.untreatedNames <- names(jg.EROnlyCountsTreated)
jg.treatedNames   <- names(jg.EROnlyCountsUntreated)

@@

g.correctionFactor<-jg.getCorrectionFactor(jg.EROnlyExperimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames
)

@@

@@

```{r,results='hide', message=FALSE, warning=FALSE}
filename<-"Rdata/066_SLX-15090_dba_human_ER_CTCF_merged.rda"
load(filename)

@@

jg.dba<-Brundle(
            dbaExperiment,
            dbaControl,
            jg.treatedCondition,
            jg.untreatedCondition,
            jg.experimentSampleSheet,
            jg.controlSampleSheet )

@@

```{r,results='hide', message=FALSE, warning=FALSE}

@@

```{r,results='hide', message=FALSE, warning=FALSE}

profile<-read.table("SLX-15090/Homer_Profile.txt",sep="\t")
profile_ER<-as.matrix(profile[,c(1,2)])
plot(profile_ER,t="l",col="red", xlab="Distance from TSS",ylab="Read Depth",main="H4K12ac Profile")
profile_Ctrl<-as.matrix(profile[,c(1,5)])
lines(profile_Ctrl,col="blue")
legend("topright",legend = c("Estrogen", "Control"), col=c("red","blue"),pch = c(19,19))
abline(v=0, lty=3)

@@

```{r,results='hide', message=FALSE, warning=FALSE}

jg.dba<-Brundle(
            dbaExperiment,
            dbaControl,
            jg.treatedCondition,
            jg.untreatedCondition,
            jg.experimentSampleSheet,
            jg.controlSampleSheet )

```{r,results='hide', message=FALSE, warning=FALSE}

>>>>

## MA plot of H2Av normalization
```{r H2AvNormalisationPlots, fig.cap="MA plots showing ER binding before and after treatment with fulvestrant including matched Dm H2Av spike-in control.} (A) Reads corrected to total aligned reads showed the same off-centre peak density as observed in Figure 1. Putative unchanged ER binding sites are within the red triangle. (B) Overlaying the MA plot combining the changes in chromatin binding of Hs ER (black) and Dm H2Av (blue). Dm peaks overlay the off-centre peak density. (C) Utilising the Dm H2Av binding events as a ground truth for 0-fold change, a linear fit to the log-fold change is generated and the fit is applied to adjust the Hs ER binding events."}
par(mfrow=c(1,3))
# Read consensus peaks
hsconsensus<-readRDS("Rdata/015_SLX-8047_hsconsensus.rds")
dmconsensus<-readRDS("Rdata/015_SLX-8047_dmconsensus.rds")

@@

Figure 2 can be viewed interactively on the [USCS track](http://genome.ucsc.edu/cgi-bin/hgTracks?hgS_doOtherUser=submit&hgS_otherUserName=andrewholding&hgS_otherUserSessionName=Brundle).

##H2av with DiffBind

```{r H2avDiffBind, results='hide', message=FALSE, warning=FALSE,   out.width = '49%', out.extra='style="float:left"'}
jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX8047_dm.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX8047_hs.csv"
filename<-"Rdata/029_SLX-8047_dba_human_drosophila.rda"

@@

source('package/brundle.R')

@@

```{r H2avDiffBindNormalised, results='hide', message=FALSE, warning=FALSE, out.width = '49%', out.extra='style="float:left"'}
  dbaExperimentAnalysisCorrected<-dba.analyze(dbaExperimentCorrected)
  dba.plotMA(dbaExperimentAnalysisCorrected)
```

@@

```{r DeSeqCTCFControlPeaks, results='hide', message=FALSE, warning=FALSE, fig.cap="Comparison of the control regions used to normalise ER analysis before and after treatment. Dots highlighted in red are significant (FDR = 0.01). The CTCF peaks used for normalisation show no significant change in the number reads before and after treatment."}

#This is head as otherwise next figure doesn't work.

filename<-"Rdata/040_SLX-14438_dba_human_ER_CTCF.rda"
load(filename)
jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)

#jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
#                                            jg.treatedNames,
#                                            jg.untreatedNames
#)

#Manuscript figure uses the above function, but to enable running of the code without the bam files we 
#estimate the Factor from diffbind object.

untreated <-
            as.numeric(row.names(dbaExperiment$samples)[dbaExperiment$samples$Condition ==
                                                            jg.untreatedCondition])
        treated <-
            as.numeric(row.names(dbaExperiment$samples)[dbaExperiment$samples$Condition ==
                                                            jg.treatedCondition])
        jg.correctionFactor <-
            sum(as.numeric(dbaExperiment$class['Reads', c(treated)])) / sum(as.numeric(dbaExperiment$class['Reads', c(untreated)]))

#Store this and use it later.

CTCFbindingPlot<-jg.controlResultsDeseq

@@

```{r LinearModel, results='hide', message='false', warning='false', fig.cap="Comparison of mean counts in CTCF peaks before and after treatment. If the samples have no systematic bias before and after treatment then the linear fit would be expected to have a gradient of 1. Here, we establish that the gradient is < 1, implying a systematic bias between samples. The read counts in the treated samples peaks are corrected (blue), removing the bias, and resulting in a new gradient of 1."}

@@

##Normalisation factors are consistent over a wide range in number of control binding sites

```{r CTCFStablility, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library("Brundle")

jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX14438_hs_CTCF_DBA.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX14438_hs_ER_DBA.csv"
jg.treatedCondition       =  "Fulvestrant"
jg.untreatedCondition     =  "none"

jg.coefficient_baseline<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
rows<-nrow(jg.controlCountsTreated)

jg.coefficient<-matrix(nrow=100,ncol=100)

for (n in 100:1)
{
    for (i in 1:100)
    {
        set.seed(i)
        sampleRows<-sample(rows, rows*(n/100) )
        jg.coefficient[n,i]<-jg.getNormalizationCoefficient(jg.controlCountsTreated[sampleRows, ],
                               jg.controlCountsUntreated[sampleRows, ])
    }
}

plot( rowMeans(jg.coefficient)/jg.coefficient_baseline*100-100,ylim=c(-2,+2),
      xlim=c(100,0),
      pch=20,col="blue",
      type="l",
      xlab="Percent of CTCF sites",
      ylab="Percent Error in Normalisation Coefficient",
      main="Stablity of Normalisation Coefficient",lwd=2)
lines( apply(t(jg.coefficient), 2, max) /jg.coefficient_baseline*100-100,col="black",lty=2,lwd=2) #max=1.956436
lines( apply(t(jg.coefficient), 2, min)/jg.coefficient_baseline*100-100, col="black",lty=2,lwd=2) #min=-1.970216
lines( apply(t(jg.coefficient), 2, quantile)['25%',]/jg.coefficient_baseline*100-100 ,col="black", lwd=2)#min=-0.4968219
lines( apply(t(jg.coefficient), 2, quantile)['75%',]/jg.coefficient_baseline*100-100, col="black",lwd=2)#max=-0.5254374
legend("topleft", legend=c("Average", "25%/75% quantile","Max/min error"),
       col=c("blue", "black","black"), lty=1:1:2, cex=0.8)

Stability of CTCF derived normalisation coefficient. Stability of the CTCF derived normalisation coefficient was analysed by sub-sampling CTCF peaks before undertaking the calculation (between 1-100% of total sites) at random. This analysis was repeated 100 times to model the variability of the result.

@@

```{r CTCF.H2avComparision, results='hide', echo=FALSE,warning=FALSE,message=FALSE,error=FALSE} 
jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX8047_dm.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX8047_consensus.csv"

@@

```{r CTCF.H2avOverlay, out.width = '50%', out.extra='style="float:left"'}
a + as.layer(b) 
```

```{r CTCF.H2avXYScatter, out.width = '50%', out.extra='style="float:left"', fig.cap="Comparison of normalisation methods using consensus peak set. (A) The analysis for the CTCF normalised (blue) and H2Av normalised (green) dataset using an ER consensus peak set of 10,000 peaks were formatted as an MA plot and overlaid. This recovered the low-fold change higher-intensity peaks that were not visible in Figure \ref{fig:ERCTCF}A and both datasets showed a similar distribution. (B) Comparison of fold-change values for individual ER binding sites between two datasets showed that the inclusion of these sites did not appear to affect the correlation (r = 0.77)."}

@@

```{r LoadDiffBind, results='hide', message=FALSE, warning=FALSE}
library(DiffBind)

```{r ClassicMethods, fig.cap="Comparison of simple normalisation strategies employed. MA plots showing the changes in ER binding after 48 hours treatment with 100 nM fulvestrant. Three simple normalisation methods were applied to this data and compared to the raw count data. (A) Raw counts. (B) Reads Per Million (RPM) reads in peaks. (C) RPM aligned reads. (D) RPM total reads. Note that the highlighted peaks remain above zero under all three standard normalisations."}

@@

```{r AnalysisPackageComparison, results='hide', message=FALSE, warning=FALSE}

@@

jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)

#Publication uses the counts from the Bam file, but to save space here we get them from the DiffBind object

#jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
#                                            jg.treatedNames,
#                                            jg.untreatedNames
#                                            )

untreated <-
            as.numeric(row.names(dbaExperiment$samples)[dbaExperiment$samples$Condition ==
                                                            jg.untreatedCondition])
        treated <-
            as.numeric(row.names(dbaExperiment$samples)[dbaExperiment$samples$Condition ==
                                                            jg.treatedCondition])
        jg.correctionFactor <-
            sum(as.numeric(dbaExperiment$class['Reads', c(treated)])) / sum(as.numeric(dbaExperiment$class['Reads', c(untreated)]))

@@

```{r ReproducbilityXYScatter, fig.cap="Correlation Plots of Replicate Experiments. (A) Scatter plots showing the correlation between the replicates with the lowest correlation value. This is provided both the control (top) and treatment (bottom) conditions. The plotted condition is highlighted with thick border in tables on the right. Colour represents density, blue = lowest, red = highest. (B) Tables showing the correlation coefficient for been each replicated. "}

@@

@@

```{r MAPlotMouseSpikeIn}

@@

```{r RelativeReadAligmentMm, fig.cap="Distribution of reads for Mm chromatin spike-in normalisation strategy. Comparison of murine chromatin between samples showed no systematic bias in the sample preparation. Bar plots (left axis) represent the fraction of total aligned reads. The dot plot represents the total aligned reads (right axis) for each sample."}
filename<-"Rdata/012_SLX-12998_nrreads.rda"
load(filename)

@@

```{r CTCFNormalisationPlots}
par(mfrow=c(1,3))

@@

```{r ERCTCFHeatmap,results='hide', message=FALSE, warning=FALSE,   out.width = '33%', out.extra='style="float:left"'}
filename<-"Rdata/015_SLX-14229_dbacounts_human.rda"
load(filename)
plot(dbacounts)

@@

##Comparison of control regions

```{r H2avControlRegions, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}

 jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX8047_dm.csv"

  ## Extract Peak set from DiffBind
jg.controlPeakset    <- jg.dbaGetPeakset(dbaControl)

#Run DeSeq on control
#jg.controlDeSeq<-jg.runDeSeq(jg.controlPeaksetDeSeq, jg.conditions,jg.controlSizeFactors)
#jg.controlResultsDeseq   = results(jg.controlDeSeq)

#Some defaults in DeSeq changed after preparing figures, the next 3 steps overides these setting to get the same figures as in the manuscript.

DESeqData<-DESeqDataSetFromMatrix(jg.controlPeaksetDeSeq, jg.conditions, ~Condition)
jg.controlDeSeq<-DESeq(DESeqData,betaPrior = TRUE)
jg.controlResultsDeseq   = results(jg.controlDeSeq, MLE=TRUE)

 jg.plotDeSeq(jg.controlResultsDeseq,
               p=0.01, 
               title.main="Fold-change in H2av binding",
               flip=T
               )
```

```{r CTCFControlRegions, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}

jg.plotDeSeq(CTCFbindingPlot,
             p=0.01, 
             title.main="Fold-change in CTCF binding",
             flip=T
)

Comparison of the control regions used to normalise ER analysis before and after treatment. Dots highlighted in red are significant (FDR = 0.01). (A) H2Av occupancy of the Drosophila genome shows no significant changes before and after treatment. (B) The CTCF peaks used for normalisation show no significant change in the number reads before and after treatment.

@@

```{r DeSEQDmNormalisation, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}

@@

```{r DeSEQCTCFNormalisation, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}

@@

##Comparison of normalisation DiffBind plots

```{r DiffBindNormalisation,results='hide', message=FALSE, warning=FALSE,   out.width = '32%', out.extra='style="float:left"'}
jg.controlSampleSheet     <- "samplesheet/samplesheet_SLX14438_hs_CTCF_DBA.csv"
jg.experimentSampleSheet  <- "samplesheet/samplesheet_SLX14438_hs_ER_DBA.csv"

@@

jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)

#Estiamte correction factor from Diffbind Object as Bams may not be avalible

untreated <-
            as.numeric(row.names(dbaExperiment$samples)[dbaExperiment$samples$Condition ==
                                                            jg.untreatedCondition])
        treated <-
            as.numeric(row.names(dbaExperiment$samples)[dbaExperiment$samples$Condition ==
                                                            jg.treatedCondition])
        jg.correctionFactor <-
            sum(as.numeric(dbaExperiment$class['Reads', c(treated)])) / sum(as.numeric(dbaExperiment$class['Reads', c(untreated)]))

#jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
#                                            jg.treatedNames,
#                                            jg.untreatedNames)

@@

```{r CrossNormalisationXYScatter, results='hide', message=FALSE, warning=FALSE,  out.width = '50%', out.extra='style="float:left"'}
filename<-"Rdata/047_SLX-14438_dba_human_ER_CTCF.rda"
load(filename)

@@

jg.treatedNamesPF<-jg.treatedNames
jg.untreatedNamesPF<-jg.untreatedNames

@@

#Get sample names for conditions
jg.treatedNames <- names(jg.ERCountsTreated)
jg.untreatedNames   <- names(jg.ERCountsUntreated)

@@

#Get sample names for conditions
jg.untreatedNames <- names(jg.EROnlyCountsUntreated)
jg.treatedNames   <- names(jg.EROnlyCountsTreated)

@@

untreated <-
            as.numeric(row.names(dbaERonlyExperiment$samples)[dbaERonlyExperiment$samples$Condition ==
                                                            jg.untreatedCondition])
        treated <-
            as.numeric(row.names(dbaERonlyExperiment$samples)[dbaERonlyExperiment$samples$Condition ==
                                                            jg.treatedCondition])
        jg.correctionFactor <-
            sum(as.numeric(dbaERonlyExperiment$class['Reads', c(treated)])) / sum(as.numeric(dbaERonlyExperiment$class['Reads', c(untreated)]))

#jg.correctionFactor<-jg.getCorrectionFactor(jg.EROnlyExperimentSampleSheet,
#                                            jg.treatedNames,
#                                            jg.untreatedNames
#)

@@

```{r CrossNormalisationBoxPlot, message=FALSE, warning=FALSE, out.width = '50%', out.extra='style="float:left"', fig.cap="Comparison of fold-change of ER binding after both xenogenic and cross-normalisation. (A) Scatter plot of fold-change as established at individual sites by each method.  Pearson's correlation between the two methods is 0.992 (3sf, p-value tending to 0). Deviation of data points from parity is a result the integer nature of read counts, nonetheless this effect is is very small as demonstrated correlation coefficient between the two datasets. (B) Box-plot showing the fold-change of ER binding before and after treatment at ∼550 ER sites proximal to CTCF binding. The mean and maximum fold-change is reduced at these sites by Parallel-Factor ChiP, but the effect is marginal."}

library(Brundle)

erChIP<-makeGRangesFromDataFrame(jg.EROnlyExperimentPeakset,keep.extra.columns=TRUE)
pfChIP<-makeGRangesFromDataFrame(jg.normalisedExperimentPeakset,keep.extra.columns=TRUE)
ctcfBED<-read.delim("SLX-14438_merged/peaks/CTCF_union.bed",header = F)
colnames(ctcfBED)<-c( "CHR" ,"START","END")
jg.controlPeaksetWide<-ctcfBED
jg.controlPeaksetWide[2]<-ctcfBED[2]-1000
jg.controlPeaksetWide[3]<-ctcfBED[3]+1000
ctcfChIP<-makeGRangesFromDataFrame(jg.controlPeaksetWide)

#jg.plotMA(jg.normalisedExperimentPeakset,jg.normalisedExperimentPeakset,jg.untreatedNamesPF,jg.treatedNamesPF,1)
#jg.plotMA(jg.EROnlyExperimentPeakset,jg.EROnlyExperimentPeakset,jg.treatedNames,jg.untreatedNames,1/jg.ERNormalisationRatio)

erChIPctcfOverlap<-subsetByOverlaps(erChIP,ctcfChIP)
pfChIPctcfOverlap<-subsetByOverlaps(pfChIP,ctcfChIP)

erChIPpfOverlap<-subsetByOverlaps(erChIPctcfOverlap,pfChIPctcfOverlap)
pfChIPerOverlap<-subsetByOverlaps(pfChIPctcfOverlap,erChIPctcfOverlap)

erChIPuntreated<-rowMeans(as.matrix(elementMetadata(erChIPpfOverlap)[,jg.untreatedNames]))
erChIPtreated<-rowMeans(as.matrix(elementMetadata(erChIPpfOverlap)[,jg.treatedNames]))
#erChIPtreated/erChIPuntreated

pfChIPuntreated<-rowMeans(as.matrix(elementMetadata(pfChIPerOverlap)[,jg.untreatedNamesPF]))
pfChIPtreated<-rowMeans(as.matrix(elementMetadata(pfChIPerOverlap)[,jg.treatedNamesPF]))
#pfChIPtreated/pfChIPuntreated

boxplot(log(erChIPtreated/erChIPuntreated),
        log(pfChIPtreated/pfChIPuntreated),
        xlab="Experiment",
        ylab="Log(Fold-Change)",
        main=paste("Fold-Change of ER binding","Proximal to CTCF",collapse="\n"),
        names=c(" Cross-Normalised"," Parrallel Factor"),
        pch=20)

```{r ERNormalisedMAPlot,results='hide', message=FALSE, warning=FALSE}

jg.experimentPeaksetNormalised<-Brundle(
            dbaExperiment,
            dbaControl,
            jg.treatedCondition,
            jg.untreatedCondition,
            jg.experimentSampleSheet,
            jg.controlSampleSheet,
            jg.noBAMs = TRUE ) #As previous examples, BAMs were used in the manuscript.

jg.dba <- DiffBind:::pv.resetCounts(dbaExperiment, jg.experimentPeaksetNormalised)

```{r ERChIPReproducibility,results='hide', message=FALSE, warning=FALSE}

## Comparision of ER binding from public datasets

Comparison ER binding from public datasets.Common peaks detected for ER Ross-Innes CS, et al. 2010; Welboren WJ, et al., 2009; Ceschin DG, et al. 2011 and our data (FDR = 0.01). Venn diagram was generated with ChIPSeqAnno.
```{r ERVennOverlap, message=FALSE, warning=FALSE}
library(ChIPpeakAnno)
library(GenomicRanges)

m1 = read.table("bed/ERoverlap/Carrol-GenesDev.bed", sep="\t")
m2 = read.table("bed/ERoverlap/ER_45minsGains_001_hg18.bed", sep="\t")
m3 = read.table("bed/ERoverlap/LinER.txt", sep="\t")
m4 = read.table("bed/ERoverlap/Welboren.bed", sep="\t")
m5 = read.table("bed/ERoverlap/NA_peaks.narrowPeak", sep="\t") #Called with macs2 against EtOH control
m5<-m5[c(1:3)]

m1.r = BED2RangedData(m1)
m2.r = BED2RangedData(m2)
m3.r = BED2RangedData(m3)
m4.r = BED2RangedData(m4)
m5.r = BED2RangedData(m5)
#mkd = makeVennDiagram(list(m1.r,m2.r,m3.r,m4.r,m5.r),
#                      NameOfPeaks=c("Ross-Innes 2010", "Guertin 2017", "Lin 2007", "Welboren 2009","Ceschin 2011"), totalTest=200000)

mkd = makeVennDiagram(list(m1.r,m2.r,m4.r,m5.r),
                      NameOfPeaks=c("Ross-Innes 2010", "Guertin 2017", "Welboren 2009","Ceschin 2011"))

@@

```{r H4K12acNormalised,results='hide', message=FALSE, warning=FALSE}
filename<-"Rdata/066_SLX-15090_dba_human_ER_CTCF_merged.rda"
load(filename)

@@

jg.experimentPeaksetNormalised<-Brundle(
            dbaExperiment,
            dbaControl,
            jg.treatedCondition,
            jg.untreatedCondition,
            jg.experimentSampleSheet,
            jg.controlSampleSheet,
            jg.noBAMs=TRUE
            ) #BAMS we used in the manuscript data, but removed to keep file size down.

jg.dba <- DiffBind:::pv.resetCounts(dbaExperiment, jg.experimentPeaksetNormalised)

@@

```{r H4K12acReproducibility,results='hide', message=FALSE, warning=FALSE}

@@

```{r H4K12acProfile,results='hide', message=FALSE, warning=FALSE}

profile<-read.table("SLX-15090/Homer_ER_Profile.txt",sep="\t")
profile_ER<-as.matrix(profile[,c(1,2)])
plot(profile_ER,t="l",col="red", xlab="Distance from ER Binding site",ylab="Read Depth",main="H4K12ac Profile")
profile_Ctrl<-as.matrix(profile[,c(1,5)])
lines(profile_Ctrl,col="blue")
legend("topright",legend = c("Estrogen", "Control"), col=c("red","blue"),pch = c(19,19))
abline(v=0, lty=3)

@@
